---
title: "Introduction to ragent"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ragent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `ragent` package provides AI agents that can perform tasks using local LLMs (Large Language Models). This vignette introduces the two main agents: the calculator agent for mathematical operations and the RAG agent for document search and question answering.

## Calculator Agent

The calculator agent combines R's computational power with natural language understanding. It can:
- Parse natural language queries into mathematical expressions
- Perform calculations safely
- Provide human-friendly explanations of results

```{r, eval = FALSE}
library(ragent)

# Simple arithmetic with explanation
agent_calculator("What is 123.434235435345 times the square root of 2?")
#> The calculation '123.434235435345 * sqrt(2)' equals approximately 174.57.
#> This multiplies a large number by the square root of 2 (≈1.4142).

# Percentage calculations
agent_calculator("What is 15% of 850?")
#> 15% of 850 equals 127.50.
#> This is calculated as 850 * 0.15.

# Mathematical functions
agent_calculator("What is the square root of 16?")
#> The square root of 16 is 4, which means 4 x 4 = 16.
```

The agent works in three steps:
1. Parses the natural language query into a mathematical expression
2. Evaluates the expression safely using R
3. Generates a human-friendly explanation of the result

You can customize the model used for explanations:
```{r, eval = FALSE}
# Use a different model for more detailed explanations
agent_calculator("What is pi times 5?", model = "deepseek-r1:8b")
#> 5 times pi equals approximately 15.71 (rounded to 2 decimal places).
#> Pi (π) is the mathematical constant representing the ratio of a circle's
#> circumference to its diameter.
```

## RAG Agent (Retrieval Augmented Generation)

The RAG agent helps answer questions by searching through a knowledge base of documents. It uses:
- Vector embeddings for semantic search
- Multi-step retrieval for comprehensive answers
- Source citations for transparency

Here's how to use the RAG agent with a simple knowledge base:

```{r}
# Create a simple knowledge base for demonstration
dir.create("kb_demo")
writeLines(
  "Office hours are Monday to Friday, 9 AM to 5 PM.
  The cafeteria is open from 8 AM to 3 PM.
  Staff meetings are held every Tuesday at 10 AM.",
  "kb_demo/office.md"
)
```

```{r, eval = FALSE}
# Ask questions about the knowledge base
agent_rag("What are the office hours?", dir = "kb_demo")
#> Based on the information in the documents:
#> The office hours are Monday to Friday, from 9 AM to 5 PM.
#> [Source: office.md]

# The agent can handle follow-up questions
agent_rag("When are staff meetings held?", dir = "kb_demo")
#> According to the documents:
#> Staff meetings are scheduled for every Tuesday at 10 AM.
#> [Source: office.md]

# Use more context chunks for detailed answers
agent_rag(
  "Tell me about all scheduled meetings and hours.",
  dir = "kb_demo",
  n_context = 5
)
#> Based on the available information:
#> - Office hours: Monday to Friday, 9 AM to 5 PM
#> - Staff meetings: Every Tuesday at 10 AM
#> - Cafeteria hours: 8 AM to 3 PM
#> [Source: office.md]
```

The RAG agent works in multiple steps:
1. Performs initial semantic search to find relevant context
2. Generates follow-up questions to gather more information
3. Combines all retrieved information to provide a comprehensive answer
4. Includes citations to source documents

## Next Steps

The `ragent` package is designed to be extensible:
- Create custom tools for specific tasks
- Combine multiple agents for complex workflows
- Use different LLM models for different needs

Check the package documentation for more advanced features and customization options.

```{r, include = FALSE}
# Clean up demo files
unlink("kb_demo", recursive = TRUE)
```
