---
title: "Playing with AI Agents in R"
format: html
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

What a ride since I've first started to [play with local LLMs](https://blog.cynkra.com/posts/2024-07-27-llama3.1/).
With various new Llama models, we got awesome small and large models that run smoothly on consumer hardware.
With deepseek R1, we got some ridicoulously good MIT-licensed models that are competitive with the best models from OpenAI.

And now, we also got a shiny fully-speced M4 Mac Mini for our office that runs all of this like a charm. It's local LLM time! 🥳

Let's explore how we can use all of this to run agents in R. I created a simple toy package to demonstrate some key concepts of AI agents in R.

## What is an agent?

Agents are chats that can use tools do acutal stuff. And R is a wonderful language to tell tools what to do.

Think about it this way:

1. You tell an LMM what you want in plain English and the tools you have available
2. The LMM figures out what tools it needs to use
3. The agent uses the tools to get the job done
4. Iterate, improve or continue with other tasks
5. Let the LMM explain what it did

## R and Ollama

For the following experiments with agents, I'm using plain R and ollama.

I know that there is the R package [ellmer](https://ellmer.tidyverse.org), and you can do many thinks that I am doint here with it.
While [ellmer](https://ellmer.tidyverse.org) is a great package, I prefer to work directly with the ollama API. In my opinion, this gives us more transparency of the agent mechanics.

My experiments with agents in R can be found in the [ragent](https://github.com/cynkra/ragent) repository on GitHub. To install the package, use:

```{r}
#| eval: false
remotes::install_github("cynkra/ragent")
```

I will use diagrams to show the conversation flow and the tools that are used.

## A simple chat

Getting started is straightforward. In `llm_chat()` I've used the `http2` package to wrap the ollama API.

```{r}
#| eval: false
library(ragent)
llm_chat("What is the capital of France?")
#> Paris is the capital of France.
```


```{mermaid}
sequenceDiagram
    autonumber
    participant User
    participant LLM

    User->>LLM: Question
    Note over LLM: No tools needed
    LLM-->>User: Direct response
```


No surprise here. It's just a chat: The user asks a question (1) and the LLM answers (2). By default, I am using the [`llama3.2:3b`](https://ollama.com/library/llama3.2) model, because it's fast and good enough for this demo. But I can change it to any other model that ollama supports. E.g,:

```{r}
#| eval: false
llm_chat("What is the capital of France?", model = "deepseek-r1:14b")
```


## Structured output

The ollama API supports structured output. We can use this to tell the LLM to return a specific format. For example:

```{r}
#| eval: false
llm_chat("What is the capital of France?", format = list(
  type = "object",
  properties = list(
    country = list(type = "string"),
    capital = list(type = "string")
  ),
  required = c("country", "capital")
))
#> $country
#> [1] "France"
#>
#> $capital
#> [1] "Paris"
```

This structured response makes it easy to process the output programmatically in R, as it will be automatically converted to a list with named elements. It is often a good idea to add a system prompt to guide the LLM to the desired output.


## A agent that can calculate

This is now a sweet setup for tinkering. We can use R to work with this output to do more complex things.

Note that LLMs are not really good at calculating. The small Llama model that I am using here messes this up:

```{r}
#| eval: false
llm_chat("What is 2.111^2.111")
#> [1] "To calculate this, we can use the formula for squaring a binomial:\n\n(a + b)^2 = a^2 + 2ab + b^2\n\nIn this case, a = 2 and b = 1.111.\n\nSo, \n\n(2 + 1.111)^2 = \n(3.111)^2 = 9.377226\n2*3.111 = 6.222442\n6.222442 + 9.377226 = 15.600668\n\nTherefore, 2.111^2.111 ≈ 15.60"
```

But we can tell the LLM to use a calculator tool to do the calculation in R. The result looks like this:

```{r}
#| eval: false
agent_calculator("What is 2.111^2.111")
#> **Calculation Result:** 2.111^2.111 = 4.84166414903285
```


```{mermaid}
sequenceDiagram
    autonumber
    participant User
    participant Agent
    participant Calculator

    User->>Agent: Math question
    Note over Agent: Analyzes question<br/>Decides to use calculator
    Agent->>Calculator: Structured calculation request
    Calculator-->>Agent: Numerical result
    Note over Agent: Formats response
    Agent-->>User: Explained calculation
```

Much better! This is an agent that:

1. Analyzes the question
2. Decides to use the calculator tool
3. Uses the calculator tool to calculate the result in R
4. Returns the response to the user

## A smarter RAG

A basic RAG system works by querying a knowledge base for relevant passages to the question first, and then using that context to answer the question.


```{mermaid}
sequenceDiagram
    autonumber
    participant User
    participant Agent
    participant RAG
    participant Docs

    User->>Agent: Question about docs
    Note over Agent: Analyzes question
    Agent->>RAG: Initial search
    RAG->>Docs: Search in directory
    Docs-->>RAG: Initial passages
    RAG-->>Agent: Initial context
    Note over Agent: Formulates follow-up<br/>questions for context
    Agent->>RAG: Additional searches
    RAG->>Docs: Search with new questions
    Docs-->>RAG: Additional passages
    RAG-->>Agent: Combined context
    Note over Agent: Synthesizes all<br/>information
    Agent-->>User: Complete response<br/>with citations
```

In our smarter RAG, we build an agent that

- Search the knowledge base for relevant passages (2)
- Ask the LLM to provide additional questions to search for (5)
- Search the knowledge base for the additional questions (7)
- Use the context to answer the question (10)


```{r}
#| eval: false
agent_rag("What is ggplot2 used for?", dir = "docs/")
#> Based on the documentation:
#> ggplot2 is a core package of the tidyverse used for visualization.
#> [Source: tidyverse.md]
```

This is just the beginning. We can build much more complex agents. For example, we could create a general agent that combines both the calculator and RAG capabilities. This agent would analyze each question and automatically choose the right approach - either searching the knowledge base or performing calculations.

## Conclusion

R is a perfect language for building AI agents. Its functional nature makes tools feel like natural building blocks that we can freely combine. The way R handles lists and data frames fits seamlessly with JSON, and the rich ecosystem provides everything we need to build powerful agents. Best of all, R's interactive workflow lets us watch our agents in action, making development feel like a conversation itself.
